---
title: "Running the dosefitr imaging pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running the dosefitr imaging pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
library(dosefitr)
```

## Overview

The `dosefitr` package provides a three-step pipeline for processing high-content live-cell imaging data acquired on the IncuCyte platform:

1. **`run_merge_channels()`** — merge multi-channel time-series TIFFs into a single file per well.
2. **`run_cellpose()`** — segment cells in each frame using Cellpose.
3. **`run_trackmate()`** — track segmented cells across frames using TrackMate via Fiji/ImageJ.

Each function is a thin R wrapper that calls a Python script bundled with the package via `system2()`. You will need a working Python environment with Cellpose installed, and a Fiji installation with the TrackMate plugin.

---

## Prerequisites

### Python environment

`dosefitr` ships a conda environment specification at `inst/python/environment.yaml` that pins all required dependencies, including `cellpose==4.0.8`, `torch==2.9.1`, and `segment-anything==1.0`. Using this file is the recommended way to set up a working Python environment.

#### Option A — micromamba (recommended)

The `prefix` field in `environment.yaml` targets `micromamba/envs/dosefitr`, so micromamba will install the environment there by default. If you do not already have micromamba, see <https://mamba.readthedocs.io/en/latest/installation/micromamba-installation.html>.

```{bash micromamba-install, engine="bash"}
# Locate the bundled environment file
ENV_FILE=$(Rscript -e 'cat(system.file("python", "environment.yml", package = "dosefitr"))')

# Create the environment (installs to micromamba/envs/dosefitr by default)
micromamba env create -f "$ENV_FILE"

# Activate and verify
micromamba activate dosefitr
python -c "import cellpose; print(cellpose.__version__)"
```

#### Option B — conda / mamba

If you are already using conda or mamba, you can create a named environment from the same file. The `prefix` field in the YAML will be ignored in favour of the `-n` flag.

```{bash conda-install, engine="bash"}
ENV_FILE=$(Rscript -e 'cat(system.file("python", "environment.yml", package = "dosefitr"))')

# conda env create -n dosefitr -f "$ENV_FILE"
# or, faster:
micromamba env create -n dosefitr -f "$ENV_FILE"

conda activate dosefitr
```

#### Finding the Python executable path

Once the environment is active, get the full path to the Python executable — you will need this for the `python` argument in each `dosefitr` function:

```{bash which-python, engine="bash"}
# micromamba
micromamba run -n dosefitr which python

# conda
conda run -n dosefitr which python
```

On most Linux/macOS systems this will be something like
`~/micromamba/envs/dosefitr/bin/python` or `~/miniconda3/envs/dosefitr/bin python`. Store it in a variable at the top of your analysis script for convenience:

```{r python-path}
python_bin <- "~/micromamba/envs/dosefitr/bin/python"   # adjust to your system
```

```r
# Check that the bundled Python scripts are accessible
system.file("python", "cellpose.py",       package = "dosefitr")
system.file("python", "merge_channels.py", package = "dosefitr")
system.file("python", "trackmate.py",      package = "dosefitr")
```

If you are using a conda or virtual environment, pass the full path to the Python executable via the `python` argument in each function (see below).

### Fiji / ImageJ

Download Fiji from <https://fiji.sc/> and note the path to the executable, e.g.:

- **Linux:** `/opt/fiji/Fiji.app/ImageJ-linux64`
- **macOS:** `/Applications/Fiji.app/Contents/MacOS/ImageJ-macosx`
- **Windows:** `C:/Fiji.app/ImageJ-win64.exe`

---

## Step 1 — Merge channels

If your IncuCyte experiment exports each fluorescence channel into a separate directory, use `run_merge_channels()` to combine them into a single multi-channel TIFF per well before segmentation.

```{r merge-basic}
run_merge_channels(
  input_dirs = c("/data/experiment1/phase",
                 "/data/experiment1/green"),
  output_dir = "/data/experiment1/merged"
)
```

For larger datasets you can parallelise across cores:

```{r merge-parallel}
run_merge_channels(
  input_dirs = c("/data/ch1", "/data/ch2", "/data/ch3"),
  output_dir = "/data/merged",
  cores      = 8
)
```

If the Python process exits with a non-zero status, `run_merge_channels()` will throw an informative error so the pipeline stops early rather than propagating bad data to downstream steps.

---

## Step 2 — Cellpose segmentation

`run_cellpose()` takes the merged (or single-channel) directory and produces `_cp_masks.tif` files alongside each input image.

### Basic usage — single channel

```{r cellpose-basic}
run_cellpose(
  input_dir = "/data/experiment1/merged"
)
```

### Specifying the segmentation channel

Use the `channels` argument to select which channel to segment on (1-based). For example, to segment on channel 2 (e.g. a nuclear marker):

```{r cellpose-channel}
run_cellpose(
  input_dir = "/data/experiment1/merged",
  channels  = 2
)
```

### Passing extra Cellpose CLI arguments

Any additional Cellpose command-line arguments can be forwarded via
`cellpose_args`. They are appended after `--` and passed verbatim to the Cellpose CLI.

```{r cellpose-args}
# Use the 'cyto2' model with a custom diameter
run_cellpose(
  input_dir     = "/data/experiment1/merged",
  channels      = 1,
  cellpose_args = c("--diameter", "30")
)
```

### Using a custom Python executable

```{r cellpose-python}
run_cellpose(
  input_dir = "/data/experiment1/merged",
  python    = "/miniconda3/envs/cellpose/bin/python"
)
```

---

## Step 3 — TrackMate cell tracking

`run_trackmate()` merges the Cellpose mask TIFFs with their source images, generates a TrackMate macro, and executes it via the Fiji CLI. Output `.xml` track files and `_spots.csv` tables are written to a `trackmate/` subdirectory inside `input_dir`.

### Basic usage

```{r trackmate-basic}
run_trackmate(
  input_dir = "/data/experiment1/merged",
  fiji_path = "Fiji.app/ImageJ-linux64"
)
```

### Tightening linking parameters

The default linking distances are conservative (200 px). For densely packed cells or slower-moving populations you will typically want to tighten these:

```{r trackmate-tight}
run_trackmate(
  input_dir                  = "/data/experiment1/merged",
  fiji_path                  = "/opt/fiji/Fiji.app/ImageJ-linux64",
  frame_linking_max_distance = 50,
  gap_closing_max_distance   = 50,
  gap_closing_max_frame_gap  = 2L
)
```

### Cell division (track splitting)

Track splitting is enabled by default, which allows TrackMate to detect cell
division events. You can tune the maximum distance at which a split is accepted:

```{r trackmate-split}
run_trackmate(
  input_dir                = "/data/experiment1/merged",
  fiji_path                = "/opt/fiji/Fiji.app/ImageJ-linux64",
  allow_track_splitting    = TRUE,
  track_splitting_max_distance = 20
)
```

To disable division detection entirely (e.g. for non-dividing populations):

```{r trackmate-nosplit}
run_trackmate(
  input_dir             = "/data/experiment1/merged",
  fiji_path             = "/opt/fiji/Fiji.app/ImageJ-linux64",
  allow_track_splitting = FALSE
)
```

### Verbose output

Set `verbose = TRUE` to print the full TrackMate configuration and the Fiji command to the console before execution — useful for debugging:

```{r trackmate-verbose}
run_trackmate(
  input_dir = "/data/experiment1/merged",
  fiji_path = "/opt/fiji/Fiji.app/ImageJ-linux64",
  verbose   = TRUE
)
```

---

## Full pipeline example

The snippet below runs the complete three-step pipeline for a two-channel experiment, using a custom Python environment and tightened tracking parameters.

```{r full-pipeline}
python_bin <- "/home/zac/miniconda3/envs/cellpose/bin/python"
fiji_bin   <- "/opt/fiji/Fiji.app/ImageJ-linux64"
base_dir   <- "/data/experiment1"

# 1. Merge phase and green channels
run_merge_channels(
  input_dirs = file.path(base_dir, c("phase", "green")),
  output_dir = file.path(base_dir, "merged"),
  cores      = 8,
  python     = python_bin
)

# 2. Segment cells on the phase channel
run_cellpose(
  input_dir = file.path(base_dir, "merged"),
  channels  = 1,
  python    = python_bin
)

# 3. Track segmented cells
run_trackmate(
  input_dir                  = file.path(base_dir, "merged"),
  fiji_path                  = fiji_bin,
  frame_linking_max_distance = 50,
  gap_closing_max_distance   = 50,
  gap_closing_max_frame_gap  = 2L,
  allow_track_splitting      = TRUE,
  track_splitting_max_distance = 20,
  verbose                    = TRUE
)
```

After `run_trackmate()` completes, the `trackmate/` subdirectory will contain one `.xml` file and one `_spots.csv` file per input image.
